\section{Introduction}

This paper describes two associative hash functions, one built on top of a
derandomized treap, and one built on top of the SeqHash~\cite{versum:ccs14}
history-independent balanced tree. Using these associative hash functions, we
can efficiently combine two hashes $\hash(a)$ and $\hash(b)$ to obtain
$\hash(a||b)$ using a merge function $\cat(\hash(a), \hash(b)) = \hash(a||b)$
in time $O(\log |a| + \log |b|)$. \footnote{Like all bounds in this paper, this
runtime bound holds with high probability.} Runtime increases because our hash
functions grow, with $|\hash(s)| = O(\log |s|)$. Our hash function is useful
for fingerprinting and is collision-resistant.

Hash functions are frequently used to summarize and authenticate data. While
powerful, existing cryptographic hash functions are not associative. Imagine
creating an archive file consisting of several large input files. Using,
for example, SHA512, to compute a hash of the concatenation of two input files 
$a$ and $b$ takes linear time $O(|b|)$. Using our associative hash functions,
computing the final hash from two existing hashes would instead take time
$O(\log |a| + \log |b|)$.

Our associative hash functions are based on Merkle hash
trees~\cite{merkle:sig-conventional}. To hash a sequence of values, we build a
tree over those values, and recursively compute a Merkle hash for the tree's
root.  A Merkle hash tree computes a hash for each node in a tree by computing
a cryptographic hash over the hashes of the node's children. Tree-based hashes
are appealing because they do not require any new cryptographic assumptions; as
long as the underlying cryptographic hash is secure, the tree-based hash is
secure.  This is an important property! One alternative, the Rabin-Karp rolling
hash is not cryptographically strong (and finding collisions is trivial).
Another alternative, the Tillich-Zemor hash function based on matrices has been
shown to exhibit collisions as well. % https://eprint.iacr.org/2009/376.pdf

\begin{figure}[htb]
\begin{center}
\begin{tikzpicture}
\Tree [ .2 1 [ .3 4 ] ]
\end{tikzpicture}
%
\begin{tikzpicture}
\Tree [ .3 [ .2 1 ] 4 ]
\end{tikzpicture}
\end{center}
\caption{Two red-black trees storing the same data inserted either from left-to-right (left tree) or from right-to-left (right tree). Because the trees have different shape, the Merkle hashes of the two trees are different and not useful for fingerprinting.}
\label{fig:redblack}
\end{figure}

No associative hash functions have yet been built using Merkle hashes because
only few underlying trees can be used. To use a hash tree as the basis of an
associative hash function, the tree must have the same shape and internal
structure for the same input sequence, no matter how the tree was built. Yet
most balanced trees can take wildly different shapes depending on how they are
constructed. For example, constructing a red-black tree by appending leafs from
left-to-right leads to a different shape than appending the same leafs from
right-to-left as seen in Figure~\ref{fig:redblack}. For a hash function to be deterministic, the tree must always
have the same shape. This requirement is formalized in the notion of
history-independence~\cite{naor:hids}. 

The challenge this paper focuses on is picking and using a history-independent
tree as an associative hash function.  By reframing the problem of building a
hash function into a problem of building and manipulating trees, we transform a
cryptography problem into a datastructures problem. We outline our approach in
\S\ref{s:overview}.  To start out, we transform a treap into an associative
hash function in \S\ref{s:treap}. However, in our setting, a treap is not
guaranteed to be balanced. To improve our design, we use the
history-independent balanced tree SeqHash. We describe SeqHash in
\S\ref{s:background}, and cover our final construction in \S\ref{s:seqhash}.
Finally, we evaluate the performance of both the treap and SeqHash-based hash
functions in \S\ref{s:eval}.

\paragraph{Statement of work.} This paper builds on VerSum~\cite{versum:ccs14},
a system in which a server stores a full SeqHash tree as an authenticated
data-structure. While I had a rough idea of building an associative hash
function on top of the SeqHash tree before the class started, all work shown in
this paper was performed in the context and after the start of 6.856. The
description of SeqHash in \S\ref{s:background} was written from scratch with the
specific goal to provide greater clarity than the original description.
 % , and I came up with the improvements to SeqHash during the course
%of the project.

%This paper extends the
%prior work on SeqHash in the followings ways as 6.856 final project: We provide
%a clear explanation of the design and motivation for the SeqHash tree. We build
%an associative hash function using the SeqHash tree. We compare the behavior of
%SeqHash with another history-independent data structure. We suggest and
%evaluate improvements to the construction of SeqHash. We analyze the runtime of

%SeqHash in theory and in practice on various input datasets. All this work was
%performed in the context of 6.856.
